A Formal Model of Checked C

### 2022-07-15 sandp23/formal.tex
update.

\begin{figure}
  \small \centering
  $\begin{array}{l}
\begin{array}{lll}
\text{Function names:}~f&
       \text{Variables:}~ x
& \text{Integers:}~n::=\mathbb{Z} 
\end{array}
\\[0.5em]

@@

\text{Mode:} & m & ::= & \cmode \mid \umode \\[0.5em]

@@

     \text{Word Type:}& \tau &::=& \tint\mid \tptr{\omega}{m}
\\[0.5em]

@@

\text{Type:}&\omega &::=& \tau \mid \tallarrayb{\bvar}{\tau}
\\[0.5em]

\text{Expression:}& e & ::= & 
\evalue{n}{\tau} \mid x \mid \emalloc{\omega} \mid\elet{x}{e}{e} \\[0.2em]
&&\mid&
\ecast{\tau}{e} \mid \edyncast{\tau}{e}\mid \ecall{f}{\overline{e}} \mid \estrlen{x} \\[0.2em]
&&\mid&
\ebinop{e}{e} \mid\estar{e}\mid\eassign{e}{e}\mid\eunchecked{e}
\\[0.2em]
&&\mid&\eif{e}{e}{e}
\end{array}
    \end{array}
  $
  \caption{\lang Syntax}
  \label{fig:checkc-syn}
\end{figure}

@@

This section describes our formal model of \checkedc, called
\lang, making precise its syntax, semantics, and type system. It also
develops \lang's meta-theory, including type soundness and the blame
theorem.

@@

There are two notions of type in \lang.  Types $\tau$ classify
word-sized values including integers and pointers, while types
$\omega$ classify multi-word values such as arrays, null-terminated
arrays, and single-word-size values.
%
Pointer types ($\tptr{\omega}{m}$) include a mode annotation ($m$)
which is either checked (\cmode) or unchecked (\umode) and a type
($\omega$) denoting valid values that can be pointed to. Array types include both the type of
elements ($\tau$) and a bound ($\bvar$) comprised of an upper and
lower bound on the size of the array ($(b_l,b_h)$). Bounds $b$ are
limited to integer literals $n$ and expressions $x + n$.
Whether an array pointer is null terminated or not is determined by annotation
$\kappa$, which is $nt$ for null-terminated arrays, and $\cdot$
otherwise (we elide the $\cdot$ when writing the type). Here is the
corresponding \checkedc syntax for these types:
\[
\begin{array}{rcl}
$\code{array_ptr<}$\tau$\code{> : count(}$n$\code{)}$
&\Leftrightarrow& \tarrayptr{0}{n}{\tau}{\cmode}
\\[0.2em]
$\code{nt_array_ptr<}$\tau$\code{> : count(}$n$\code{)}$
&\Leftrightarrow& \tntarrayptr{0}{n}{\tau}{\cmode}
\end{array}
\]
As a convention we write $\tptr{\tarrayb{b}{\tau}}{\cmode}$ to mean
$\tptr{\tarray{0}{b}{\tau}}{\cmode}$, so the above examples could
be rewritten $\tptr{\tarrayb{n}{\tau}}{\cmode}$ and
$\tptr{\tntarrayb{n}{\tau}}{\cmode}$, respectively.

@@

\begin{figure}
{\small
$    \begin{array}{l}
    \begin{array}{lll}
\mu & ::= & \evalue{n}{\tau} \mid \bot\\
e & ::= & \ldots \mid \ret{x}{\mu}{e}\\
r & ::= & e \mid \enull \mid \ebounds\\
E &::=& \Box \mid \elet{x}{E}{e} \mid \ecall{f}{\overline{E}}\mid\ecast{\tau}{E} \mid \edyncast{\tau}{E} \\[0.2em]
&&\mid \ret{x}{\evalue{n}{\tau}}{E}\mid\ebinop{E}{e} \mid \ebinop{\evalue{n}{\tau}}{E}\mid\estar{E}\mid\eassign{E}{e} \\[0.2em]
&&\mid\eassign{\evalue{n}{\tau}}{E}\mid\eunchecked{E}\mid\eif{E}{e}{e}\\[0.2em]
\overline{E} &::=& E \mid  \evalue{n}{\tau},\overline{E} \mid \overline{E}, e
\end{array}
\\ \\
    \end{array} 
$
  \begin{mathpar}
    \inferrule{ m=\mode(E) \\
      e=E[e'] \\
      (\varphi,\heap,e') \longrightarrow (\varphi',\heap',e'')}
    {(\varphi,\heap,e)\longrightarrow_{m} (\varphi',\heap',E[e''])}

@@

    \inferrule[S-Ret]{}{(\varphi,\heap,\ret{x}{\mu}{\evalue{n}{\tau}}) \longrightarrow (\varphi[x\mapsto \mu],\heap,\evalue{n}{\tau})}

@@

      \inferrule[T-LetStr]
                {\Gamma(y) = \tntarrayptr{b_l}{b_h}{\tau_a}{\cmode} \\ x \not\in \fv(\tau) \\\\
                 \Gamma[x\mapsto \tint,y\mapsto [\tntarrayptr{b_l}{x}{\tau_a}{\cmode}];\Theta[x\mapsto \texttt{ge}\_0] \vdash_m e : \tau}
                {\Gamma;\Theta \vdash_m \elet{x}{\estrlen{y}}{e} : \tau}

>>>>

\begin{figure}
  \small \centering
  \[ \hspace*{-1em}
\begin{array}{l}
\begin{array}{ll}
       \text{Variables:}~ x
& \text{Integers:}~n::=\mathbb{Z} 
\end{array}
\\[0.5em]

@@

\text{Context Mode:} & m & ::= & \cmode \mid \umode \\[0.5em]

\text{Pointer Mode:} & \xi & ::= & m \mid \tmode \\[0.5em]

@@

     \text{Word Type:}& \tau &::=& \tint\mid \tptr{\omega}{ \xi}
\\[0.5em]

@@

\text{Type:}&\omega &::=& \tau \mid \tallarrayb{\bvar}{\tau} \mid \tfun{\overline{x}}{\overline{\tau}}{\tau}
\\[0.5em]

\text{Expression:}& e & ::= & 
\evalue{n}{\tau} \mid x \mid \ecast{\tau}{e} \mid \edyncast{\tau}{e} \mid \ebinop{e}{e} \\[0.2em]
&&\mid& \emalloc{\omega} \mid \estrlen{x} \mid\estar{e}\mid\eassign{e}{e}  \\[0.2em]
&&\mid& \elet{x}{e}{e} \mid \eif{e}{e}{e}
\\[0.2em]
&&\mid&  \ecall{e}{\overline{e}} \mid
\echecked{\overline{x}}{\tau}{e}
\\[0.2em]
&&\mid&\eunchecked{\overline{x}}{e}

\end{array}
    \end{array}
  \]
  \caption{\lang Syntax}
  \label{fig:checkc-syn}
\end{figure}

@@

This section describes the formal model of \systemname, called
\lang, making precise its syntax, semantics, and type system. It also
develops \lang's meta-theory, including type soundness, the non-exposure, and the non-crashing
theorem.

@@

There are two notions of type in \lang.  Types $\tau$ classify
word-sized values including integers and pointers, while types
$\omega$ classify multi-word values such as arrays, null-terminated
arrays, functions, and single-word-size values.
%
Pointer types ($\tptr{\omega}{m}$) include a mode annotation ($m$)
which is either checked (\cmode) or unchecked (\umode) and a type
($\omega$) denoting valid values that can be pointed to. Array types include both the type of
elements ($\tau$) and a bound ($\bvar$) comprised of an upper and
lower bound on the size of the array ($(b_l,b_h)$). Bounds $b$ are
limited to integer literals $n$ and expressions $x + n$.
Whether an array pointer is null terminated or not is determined by annotation
$\kappa$, which is $nt$ for null-terminated arrays, and $\cdot$
otherwise (we elide the $\cdot$ when writing the type).
\systemname function types ($\tfun{\overline{x}}{\overline{\tau}}{\tau}$)
reflect its dependent function declrations,
where $\overline{x}$ represents
a list of \tint type variables in a dependent function header,
which bind all bounds appearing in $\overline{\tau}$ and $\tau$.
We have a well-formed requirement for a function type,
such that all variables in $\overline{\tau}$ and $\tau$ are bounded by $\overline{x}$.
Here is the
corresponding \checkedc syntax for these types:
\[\hspace*{-0.5em}
\begin{array}{l}
\begin{array}{rcl}
$\code{array_tptr<}$\tau$\code{> : count(}$n$\code{)}$
&\Leftrightarrow& \tarrayptr{0}{n}{\tau}{\tmode}
\\[0.2em]
$\code{nt_array_ptr<}$\tau$\code{> : count(}$n$\code{)}$
&\Leftrightarrow& \tntarrayptr{0}{n}{\tau}{\cmode}
\end{array}
\\[0.2em]
$\code{tptr<(int)(nt_array_tptr<}$\tau$\code{> : count(}$n$\code{),}$\\
\qquad\qquad$\code{nt_array_tptr<}$\tau$\code{>)>: count(}$n$\code{))>}$
\\[0.2em]
\Leftrightarrow\;\; $\tptr{(\tfun{n}{ \tntarrayptr{0}{n}{\tau}{\tmode} * \tntarrayptr{0}{n}{\tau}{\tmode}}{\tint})}{\tmode}$
\end{array}
\]
As a convention we write $\tptr{\tarrayb{b}{\tau}}{\cmode}$ to mean
$\tptr{\tarray{0}{b}{\tau}}{\cmode}$, so the above examples could
be rewritten $\tptr{\tarrayb{n}{\tau}}{\cmode}$ and
$\tptr{\tntarrayb{n}{\tau}}{\cmode}$, respectively.

@@

\begin{figure}
{\small
$    \begin{array}{l}
    \begin{array}{lll}
\mu & ::= & \evalue{n}{\tau} \mid \bot\\
e & ::= & \ldots \mid \ret{x}{\mu}{e}\\
r & ::= & e \mid \enull \mid \ebounds\\
E &::=& \Box \mid \elet{x}{E}{e} \mid \ecall{f}{\overline{E}}\mid\ecast{\tau}{E} \mid \edyncast{\tau}{E} \\[0.2em]
&& \mid \ebinop{\evalue{n}{\tau}}{E}\mid\estar{E}\mid\eassign{E}{e} \\[0.2em]
&&\mid\eassign{\evalue{n}{\tau}}{E}\mid\eunchecked{E}\mid\eif{E}{e}{e}\\[0.2em]
\overline{E} &::=& E \mid  \evalue{n}{\tau},\overline{E} \mid \overline{E}, e
\end{array}
\\ \\
    \end{array} 
$
  \begin{mathpar}
    \inferrule{ m=\mode(E) \\
      e=E[e'] \\
      (\varphi,\heap,e') \longrightarrow (\varphi',\heap',e'')}
    {(\varphi,\heap,e)\longrightarrow_{m} (\varphi',\heap',E[e''])}

@@

    \inferrule[S-RetCon]{\varphi(x)=\mu\\ (\varphi[x\mapsto \mu'],\heap,e) \longrightarrow (\varphi',\heap',e')}{(\varphi,\heap,\ret{x}{\mu'}{e}) \longrightarrow (\varphi'[x\mapsto \mu],\heap',\ret{x}{\varphi'(x)}{e'})}

    \inferrule[S-RetEnd]{}{(\varphi,\heap,\ret{x}{\mu}{\evalue{n}{\tau}}) \longrightarrow (\varphi,\heap,\evalue{n}{\tau})}

@@

      \inferrule[T-LetStr]
                {\Gamma(y) = \tntarrayptr{b_l}{b_h}{\tau_a}{\cmode} \\\\
                 \Gamma[x\mapsto \tint,y\mapsto [\tntarrayptr{b_l}{x}{\tau_a}{\cmode}];\Theta[x\mapsto \texttt{ge}\_0] \vdash_m e : \tau}
                {\Gamma;\Theta \vdash_m \elet{x}{\estrlen{y}}{e} : \tau[b_h / x]}
